---
title: "Querying the WsprDaemon Timescale Database using R and SQL"
author: "Andi Fugard M0INF (almost@gmail.com, @[inductivestep](https://twitter.com/InductiveStep))"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This is a collection of examples showing how to query the [WsprDaemon Timescale Database](http://wsprdaemon.org/timebasedb.html) using R and SQL and visualise the results using the tidyverse.


## Setup packages

```{r}
library(RPostgres)
library(DBI)
library(tidyverse)
library(DT)
library(knitr)
library(fuzzyjoin)
library(tmap)
library(sf)
data(World)
dt_options <- list(pageLength=7, scrollX = "400px")
```


## Connect to the database

The WSPR database uses [PostgreSQL](https://www.postgresql.org/) which is supported in R by the `RPostgres` package.

Here are the settings required to connect -- available [over here](http://wsprdaemon.org/ewExternalFiles/Timescale_wsprdaemon_database_queries_and_APIs_V2.pdf). (Normally you shouldn't hardcode login details like this, but there is only one login and it's public.)

```{r}
db_name <- "wsprnet"
db_host <- "logs2.wsprdaemon.org"  
db_user <- "wdread"  
db_password <- "JTWSPR2008"
db_con <- dbConnect(RPostgres::Postgres(),
                    dbname = db_name,
                    host = db_host,
                    user = db_user,
                    password = db_password)  
```

Check if it worked:

```{r}
dbListTables(db_con) 
```
Yes!

Let's try an example based on SQL from the WsprDaemon Timescale Databases Notes ([Version 2.0 November 2020](http://wsprdaemon.org/ewExternalFiles/Timescale_wsprdaemon_database_queries_and_APIs_V2.pdf)):

```{r}
res <- dbSendQuery(db_con,
  "SELECT * FROM spots
   ORDER BY wd_time DESC
   LIMIT 5;")
```

```{r}
dbFetch(res) %>%
  datatable(options = dt_options)
```

Yes -- works.

```{r}
dbClearResult(res)
```


## Grab a day's data

### Transmissions 

Let's try one day's worth of data -- reception reports of M0INF. 

There's a SQL quirk below. Variable names which aren't all in lower case (such as `CallSign`) have to be referenced using double quotation marks; the quotes are escaped using the backslash. String literals like a callsign or date use single quotes.

```{r}
one_day <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE \"CallSign\" = 'M0INF'
         AND wd_time >= '2021-01-03T00:00:00Z'
         AND wd_time < '2021-01-04T00:00:00Z';")
```

This fetches the data to a data frame:

```{r}
the_dat <- dbFetch(one_day)
dbClearResult(one_day)
the_dat %>%
  datatable(options = dt_options)
```

Now plot:

```{r fig.height=4, fig.width=6, dpi = 300}
the_dat %>%
  ggplot(aes(wd_time, distance)) +
  geom_point() +
  geom_smooth(method = "gam",
              formula = y ~ s(x, k = 20)) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports of M0INF",
       subtitle = "3 Jan 2021, 20m band") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H")
```


### Both transmissions and reception reports

Let's do the same again; however, now with reception reports both by and of M0INF.

```{r}
one_day_send_receive <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE (\"CallSign\" = 'M0INF' OR \"Reporter\" = 'M0INF')
   AND wd_time >= '2021-01-03T00:00:00Z'
   AND wd_time <  '2021-01-04T00:00:00Z';")
```

```{r}
the_dat_send_rec <- dbFetch(one_day_send_receive)
dbClearResult(one_day_send_receive)
```

```{r}
the_dat_send_rec <- the_dat_send_rec %>%
  mutate(Direction = ifelse(Reporter == "M0INF",
                            "Rx by M0INF",
                            "Tx by M0INF"))
```

The graph is a bit crowded so I put the y-axis on the $\log_{10}$ scale.

```{r fig.height=5, fig.width=6, dpi = 300}
the_dat_send_rec %>%
  ggplot(aes(wd_time, distance, colour = Direction)) +
  geom_point(alpha = 0.5, size = 1) +
  stat_smooth(geom="line",
              alpha = 0.5, size = 0.8,
              method = "gam",
              formula = y ~ s(x, k = 25)) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports",
       subtitle = "3 Jan 2021, 20m band") +
  scale_y_continuous(trans = "log10") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H") +
  theme(legend.position="bottom")
```

Which of those reports are over 5000 km?

```{r}
the_dat_send_rec %>%
  filter(distance > 5000) %>%
  group_by(CallSign, Reporter, distance) %>%
  summarise(Spots = n()) %>%
  arrange(desc(Spots)) %>%
  kable()
```

And which are under 100 km?

```{r}
the_dat_send_rec %>%
  filter(distance < 100) %>%
  group_by(CallSign, Reporter, distance) %>%
  summarise(Spots = n()) %>%
  arrange(desc(Spots)) %>%
  kable()
```



## Try two days (now there's more data)


```{r}
two_day_send_receive <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE (\"CallSign\" = 'M0INF' OR \"Reporter\" = 'M0INF')
   AND wd_time >= '2021-01-03T00:00:00Z'
   AND wd_time <  '2021-01-05T00:00:00Z';")
```

```{r}
two_dat_send_rec <- dbFetch(two_day_send_receive)
dbClearResult(two_day_send_receive)
```

```{r}
two_dat_send_rec <- two_dat_send_rec %>%
  mutate(Direction = ifelse(Reporter == "M0INF",
                            "Rx by M0INF",
                            "Tx by M0INF"))
```


```{r fig.height=5, fig.width=6, dpi = 300}
two_dat_send_rec %>%
  ggplot(aes(wd_time, distance, colour = Direction)) +
  geom_point(alpha = 0.5, size = 1) +
  stat_smooth(geom="line",
              alpha = 0.5, size = 0.8,
              method = "gam",
              formula = y ~ s(x, k = 25)) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports",
       subtitle = "3-4 Jan 2021, 20m band") +
  scale_y_continuous(trans = "log10") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H") +
  theme(legend.position="bottom")
```

The red curve doesn't make sense since there is missing data which should essentially be zero. For now, let's remove it...

```{r fig.height=5, fig.width=6, dpi = 300}
two_dat_send_rec %>%
  ggplot(aes(wd_time, distance, colour = Direction)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports",
       subtitle = "3-4 Jan 2021, 20m band") +
  scale_y_continuous(trans = "log10") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H") +
  theme(legend.position="bottom")
```

### Again on the 40 m band

```{r}
day_40m_send_receive <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE (\"CallSign\" = 'M0INF' OR \"Reporter\" = 'M0INF')
   AND wd_time >= '2021-01-05T00:00:00Z'
   AND wd_time <  '2021-01-06T00:00:00Z';")
```


```{r}
day_40m_send_receive_dat <- dbFetch(day_40m_send_receive)
dbClearResult(day_40m_send_receive)
```


```{r}
day_40m_send_receive_dat <- day_40m_send_receive_dat %>%
  mutate(Direction = ifelse(Reporter == "M0INF",
                            "Rx by M0INF",
                            "Tx by M0INF"))
```


```{r fig.height=5, fig.width=6, dpi = 300}
day_40m_send_receive_dat %>%
  ggplot(aes(wd_time, distance, colour = Direction)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports",
       subtitle = "5 Jan 2021, 40m band") +
  scale_y_continuous(trans = "log10") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H") +
  theme(legend.position="bottom")
```


## Track signal strength by reporter over time...

First, look at data where there's more than one report.

```{r}
dat_sent_many <- the_dat %>%
  group_by(Reporter) %>%
  summarise(spots = n()) %>%
  arrange(desc(spots)) %>%
  slice_head(n = 7)
dat_sent_many %>%
  kable()
```



```{r fig.height=3, fig.width=6, dpi = 300}
the_dat %>%
  mutate(ReportDist = paste0(Reporter, " (", distance, "km)")) %>%
  filter(Reporter %in% unique(dat_sent_many$Reporter)) %>%
  ggplot(aes(wd_time, dB, colour = ReportDist)) +
  geom_smooth(span = 1, se = FALSE) +
  labs(x = "Time", y = "Signal report (dB)",
       title = "WSPR reports (20m) – signal strength",
       colour = "Reporter") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H")
```

It's showing something... the red curve is straightish since it's not relying on the ionosphere. The others do use ionospheric propagation so vary as conditions change.

Try again for latest 40m data:

```{r}
one_day_40m <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE \"CallSign\" = 'M0INF'
         AND wd_time >= '2021-01-05T00:00:00Z'
         AND wd_time < '2021-01-06T00:00:00Z';")
```

```{r}
dat_40m <- dbFetch(one_day_40m)
dbClearResult(one_day_40m)
```

```{r}
dat_sent_many_40 <- dat_40m %>%
  group_by(Reporter) %>%
  summarise(spots = n()) %>%
  arrange(desc(spots)) %>%
  #filter(spots >= 5)
  slice_head(n = 7)
dat_sent_many_40 %>%
  kable()
```


```{r fig.height=4, fig.width=6, dpi = 300}
dat_40m %>%
  mutate(ReportDist = paste0(Reporter, " (", distance, "km)")) %>%
  filter(Reporter %in% unique(dat_sent_many_40$Reporter)) %>%
  ggplot(aes(wd_time, dB, colour = ReportDist)) +
    geom_smooth(span = 2, se = FALSE, size = 0.8) +
    geom_point(size = 0.6, alpha = 0.5) +
    labs(x = "Time", y = "dB",
         title = "WSPR reports (40m) – signal strength",
         subtitle = "A selection of reporters on 5 Jan 2021",
         colour = "Reporter") +
    scale_x_datetime(date_breaks = "2 hour", date_labels = "%H")
```


## Last hour sent/received from IO91


### Jitter plot

Here are the WSPR frequencies, for labeling the plot in a moment.

```{r}
wspr_bands <- tibble(
  MHz = c(
    0.136,
    0.4742,
    1.8366,
    3.5686,
    5.2872,
    5364.7,
    7.0386,
    10.1387,
    14.0956,
    18.1046,
    21.0946,
    24.9246,
    28.1246,
    50.293,
    70.091,
    144.489,
    432.300,
    1296.500
  ),
  m = 299.792458 / MHz,
  wavelength_text = ifelse(m >= 1,
                           paste(round(m, 0), "m"),
                           paste(round(m*100, 0), "cm"))
) %>%
  arrange(MHz)

wspr_bands %>% kable()
```

Grab the data:

```{r}
last_whispers <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE (\"ReporterGrid\" LIKE 'IO91%'
          OR \"Grid\" LIKE 'IO91%')
   AND wd_time > NOW() - INTERVAL '1 hour';")
```


```{r}
the_dat <- dbFetch(last_whispers)
dbClearResult(last_whispers)
```


```{r}
nrow(the_dat)
```

Group into WSPR bands. This uses the `fuzzyjoin` package which matches by distance.

```{r}
joined_dat <- the_dat %>%
  difference_left_join(wspr_bands,
                       by = "MHz",
                       max_dist = 0.2,
                       distance_col = "dist_band_start") %>%
  mutate(MHz.y = as.factor(MHz.y))
```


Check how many kHz off the match is (also look for NAs):

```{r}
summary(joined_dat$dist_band_start*1000)
```


Plot:

```{r fig.height=4, fig.width=6, dpi = 300}
joined_dat %>%
  ggplot(aes(distance, MHz.y, colour = MHz.y)) +
  geom_jitter(height = 0.05, alpha = 1/3, size = 1) +
  xlab("Distance (km)") +
  ylab("Freq (MHz)") +
  labs(title = "WSPR spots IO91 (sent or received)",
       subtitle = paste0("From ", min(the_dat$wd_time),
                         " to ", max(the_dat$wd_time))) + 
  theme(legend.position = "none")
```


Try again, but this time separating the transmissions and receptions in IO91.

```{r}
joined_dat <- joined_dat %>%
  mutate(direction = ifelse(grepl(
    pattern = "IO91",
    x = Grid,
    ignore.case = TRUE
  ), "Sent from IO91", "Received in IO91"))
```


```{r fig.height=4, fig.width=7, dpi = 300}
joined_dat %>%
  ggplot(aes(distance, MHz.y, colour = direction)) +
  geom_point(position = position_jitterdodge(jitter.width = .1,
                                             jitter.height = 0,
                                             dodge.width = 0.4),
             
             alpha = 1/3, size = 1) +
  xlab("Distance (km)") +
  ylab("Freq (MHz)") +
  labs(title = "WSPR spots IO91 (sent or received)",
       subtitle = paste0("From ", min(the_dat$wd_time),
                         " to ", max(the_dat$wd_time)),
       colour = "Direction")
```


### Polar plot


Let's do the same again but also showing the direction, for transmissions from IO91 only.

```{r}
tx_only <- joined_dat %>%
  filter(grepl(pattern = "IO91", x = Grid, ignore.case = TRUE))
```

```{r fig.height=8, fig.width=8, dpi = 300}
linear_scale_polar <- tx_only %>%
  ggplot(aes(azimuth, distance, colour = MHz.y)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "WSPR spots IO91",
    subtitle = paste0("From ", min(the_dat$wd_time),
                      " to ", max(the_dat$wd_time)),
    colour = "Freq (MHz)",
    x = NULL,
    y = "Distance"
  ) +
  scale_x_continuous(
    limits = c(0, 360),
    breaks = seq(0, 315, by = 45),
    minor_breaks = seq(0, 360, by = 15)
  ) +
  coord_polar()

linear_scale_polar
```

```{r fig.height=8, fig.width=8, dpi = 300}
linear_scale_polar +
  scale_y_continuous(trans = "log10")
```

```{r}
tx_only %>%
  filter(distance > 8000) %>%
  select(CallSign, Reporter, distance, azimuth, MHz.x) %>%
  kable()
```

## A bigger one - 48 hours of spots on all frequencies to/from IO91

```{r cache=TRUE}
IO91_48hrs <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE (\"ReporterGrid\" LIKE 'IO91%'
          OR \"Grid\" LIKE 'IO91%')
         AND wd_time >= '2021-01-05T00:00:00Z'
         AND wd_time < '2021-01-07T00:00:00Z';")
```

```{r cache=TRUE}
IO91_48hrs_dat <- dbFetch(IO91_48hrs)
dbClearResult(IO91_48hrs)
nrow(IO91_48hrs_dat)
```
There were `r nrow(IO91_48hrs_dat)` rows of data.


```{r}
IO91_48hrs_dat_bands <- IO91_48hrs_dat %>%
  difference_left_join(wspr_bands,
                       by = "MHz",
                       max_dist = 0.1,
                       distance_col = "dist_band_start") %>%
  mutate(
    MHz.y = as.factor(MHz.y),
    direction = ifelse(
      grepl(
        pattern = "IO91",
        x = Grid,
        ignore.case = TRUE
      ),
      "From IO91",
      "To IO91"
    )
  )
```

Here are spots which couldn't be classified in a band:

```{r}
IO91_48hrs_dat_bands %>%
  filter(is.na(MHz.y)) %>%
  select(CallSign, Reporter, MHz.x, MHz.y, dist_band_start) %>%
  datatable(options = dt_options)
```


```{r}
summary(IO91_48hrs_dat_bands$dist_band_start*1000)
```



```{r fig.height=5, fig.width=7, dpi = 300}
IO91_48hrs_dat_bands %>%
  filter(!is.na(MHz.y)) %>%
  ggplot(aes(x = distance, fill = MHz.y)) +
  geom_histogram(aes(y=..density..), binwidth = 500, position = "identity", alpha = 0.5) +
  facet_grid(rows = vars(direction)) +
  labs(x = "Distance (km)", y = "Mass", fill = "Freq (MHz)")
```



```{r}
date_format <- "%d %b %Y (%H:%m)"
```


```{r fig.height=20, fig.width=7, dpi = 300}
IO91_48hrs_dat_bands %>%
  filter(!is.na(MHz.y)) %>%
  ggplot(aes(x = distance, fill = MHz.y)) +
  geom_histogram(
    #aes(y = ..density..),
    binwidth = 200,
    position = "identity",
  ) +
  facet_grid(cols = vars(direction), rows = vars(MHz.y)) +
  labs(
    x = "Distance (km)",
    y = "Mass",
    fill = "Freq (MHz)",
    title = "WSPR reports to/from IO91 by band",
    subtitle = paste0(format(
      min(IO91_48hrs_dat_bands$wd_time), date_format
    ), " to ",
    format(
      max(IO91_48hrs_dat_bands$wd_time), date_format
    ))
  ) +
  theme(legend.position = "none")
```

### Polar plots

```{r fig.height=40, fig.width=7, dpi = 300}
IO91_48hrs_dat_bands %>%
  filter(!is.na(MHz.y) & direction == "From IO91") %>%
  ggplot(aes(azimuth, y = distance)) +
  geom_point() +
  facet_grid(rows = vars(MHz.y)) +
  labs(
    x = "Distance (km)",
    y = "Mass",
    fill = "Freq (MHz)",
    title = "WSPR reports from IO91 by band",
    subtitle = paste0(format(
      min(IO91_48hrs_dat_bands$wd_time), date_format
    ), " to ",
    format(
      max(IO91_48hrs_dat_bands$wd_time), date_format
    ))
  ) +
  scale_x_continuous(
    limits = c(0, 360),
    breaks = seq(0, 315, by = 45),
    minor_breaks = seq(0, 360, by = 15)
  ) +
  coord_polar()
```

### Plot on maps

This needs to be tweaked but is a start. The 4326 magic number refers to the coordinate reference system. This [Wikipedia article](https://en.wikipedia.org/wiki/World_Geodetic_System) is as good a place as any to start... The `st_as_sf` function transforms longitude and lattitude coordinates to a system that the map plot can use.

```{r}
IO91_48hrs_dat_bands_sf <- IO91_48hrs_dat_bands %>%
  st_as_sf(coords = c("wd_rx_lon", "wd_rx_lat"),
           crs = 4326)
```

The following steps takes up to two minutes on my laptop, so I have cached the result in the markdown. Timing code added here to check that it worked.

```{r}
t1 <- Sys.time()
```

```{r fig.height=40, fig.width=7, dpi = 300, cache=TRUE}
IO91_48hrs_dat_bands_sf %>%
  filter(!is.na(MHz.y)) %>%
  ggplot() +
    geom_sf(data = World) +
    geom_sf(aes(colour = MHz.y)) +
    facet_grid(rows = vars(MHz.y)) +
    theme_classic() +
    labs(
      colour = "Freq (MHz)",
      title = "WSPR reports from IO91 by band",
      subtitle = paste0(format(
        min(IO91_48hrs_dat_bands$wd_time), date_format
      ), " to ",
      format(
        max(IO91_48hrs_dat_bands$wd_time), date_format
      ))
    ) +
    theme(legend.position = "none")
```


```{r}
t2 <- Sys.time()
t2 - t1
```

If the caching worked then that timer should be a few seconds.


## Try fitting a mixture model

This section seemed like a good idea at the time...


```{r}
library(mixtools)
```


First summarise the data to work out which band to play with:

```{r}
IO91_48hrs_dat_bands %>%
  filter(direction == "From IO91") %>%
  group_by(MHz.y) %>%
  summarise(spots = n(),
            dist_p25 = quantile(distance,p=.25),
            dist_p75 = quantile(distance,p=.75)) %>%
  kable()
```


I'm going to go for the 20m band since there are loads of observations and the signals carry some distance. (I should have gone for unique reports to count, but onwards...)

```{r}
dat_for_clust <- IO91_48hrs_dat_bands %>%
  filter(MHz.y == "14.0956" & direction == "From IO91")
nrow(dat_for_clust)
```

This is a bit silly since to get it work I have had to provide a lot of clues based on visual inspection... I was going to constrain the means of clusters to be multiples of the second to try to model skip. It didn't work, but I have left the code in commented out.

```{r}
cluster_n = 6
start_means <- c(0, 1500, 6000, 10000, 12000, 17000)
#means <- c(NA, paste0(1:(cluster_n-1), "n"))
sds   <- c(NA, rep("b", cluster_n-1))


clust_res <- normalmixEM(dat_for_clust$distance,
                         mu = start_means,
                         #mean.constr = means,
                         sd.constr = sds,
                         k = cluster_n)
summary(clust_res)
```

```{r}
plot(clust_res, which = 2, breaks = 100)
```


```{r cluster-hist, fig.height=3, fig.width=6, dpi = 300}
clusters <- clust_res$posterior %>%
  as.data.frame() %>%
  rowwise() %>%
  mutate(cluster = which.max(c_across()))

dat_for_clust$cluster <- ordered(clusters$cluster)

dat_for_clust %>%
  ggplot(aes(x = distance, fill = cluster)) +
  geom_histogram(alpha = 0.5,
                 position = "identity",
                 binwidth = 500) +
  labs(x = "Distance (km)",
       y = "Count",
       fill = "Cluster") +
  theme(legend.position = "bottom")
```



```{r clust_means}
clust_means <- dat_for_clust %>%
  group_by(cluster) %>%
  summarise(clust_mean_dist = mean(distance))
```




```{r}
dat_for_clust_sf <- dat_for_clust %>%
  left_join(clust_means) %>%
  st_as_sf(coords = c("wd_rx_lon", "wd_rx_lat"),
           crs = 4326)
```



```{r}
dat_for_clust_sf %>%
  ggplot() +
    geom_sf(data = World) +
    geom_sf(aes(colour = cluster), size = 1) +
    theme_classic() +
    labs(
      colour = "Cluster",
      title = "WSPR reports from IO91 (clusters)",
      subtitle = paste0(format(
        min(IO91_48hrs_dat_bands$wd_time), date_format
      ), " to ",
      format(
        max(IO91_48hrs_dat_bands$wd_time), date_format
      ))
    ) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size=3)))
```



```{r dpi = 300}
dat_for_clust_sf %>%
  ggplot(aes(x = wd_time, y = dB, colour = ordered(round(clust_mean_dist,0)))) +
  #geom_smooth(span = 3, se = FALSE, size = 2) 
  geom_jitter(size = 1, alpha = 0.3) +
  labs(colour = "Mean cluster dist (km)", x = NULL) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(override.aes = list(size=3)))
```


## db plots again - the easy way...

(To be completed...)

