---
title: "Querying the WsprDaemon Timescale Database using R and SQL"
author: "Andi Fugard M0INF (almost@gmail.com, @[inductivestep](https://twitter.com/InductiveStep))"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This short Markdown file shows how to query the [WsprDaemon Timescale Database](http://wsprdaemon.org/timebasedb.html) using R and SQL.


## Setup packages

```{r}
library(RPostgres)
library(DBI)
library(tidyverse)
library(DT)
library(knitr)
dt_options <- list(pageLength=7, scrollX = "400px")
```


## Connect to the database

The settings are available [over here](http://wsprdaemon.org/ewExternalFiles/Timescale_wsprdaemon_database_queries_and_APIs_V2.pdf).

```{r}
db_name <- "wsprnet"
db_host <- "logs2.wsprdaemon.org"  
db_user <- "wdread"  
db_password <- "JTWSPR2008"
db_con <- dbConnect(RPostgres::Postgres(),
                    dbname = db_name,
                    host = db_host,
                    user = db_user,
                    password = db_password)  
```

Check if it worked:

```{r}
dbListTables(db_con) 
```
Yes!

Let's try an example based on SQL from the WsprDaemon Timescale Databases Notes ([Version 2.0 November 2020](http://wsprdaemon.org/ewExternalFiles/Timescale_wsprdaemon_database_queries_and_APIs_V2.pdf)):

```{r}
res <- dbSendQuery(db_con,
  "SELECT * FROM spots
   ORDER BY wd_time DESC
   LIMIT 20;")
```

```{r}
dbFetch(res) %>%
  datatable(options = dt_options)
```

Yes -- works.

```{r}
dbClearResult(res)
```


## Grab a time range

Let's try one day's worth of data -- reception reports of M0INF. There's a SQL quirk below. Variable names which aren't all in lower case (such as `CallSign`) have to be referenced using double quotation marks; the quotes are escaped using the backslash. String literals like a callsign use single quotes.

```{r}
one_day <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE \"CallSign\" = 'M0INF'
         AND wd_time >= '2021-01-03T00:00:00Z'
         AND wd_time < '2021-01-04T00:00:00Z';")
```

This fetches the data to a data frame:

```{r}
the_dat <- dbFetch(one_day)
dbClearResult(one_day)
the_dat %>%
  datatable(options = dt_options)
```

Now plot:

```{r fig.height=4, fig.width=6, dpi = 300}
the_dat %>%
  ggplot(aes(wd_time, distance)) +
  geom_point() +
  geom_smooth(method = "gam",
              formula = y ~ s(x, k = 20)) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports of M0INF",
       subtitle = "3 Jan 2021, 20m band") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H")
```

Let's do the same again; however, now with reception reports both by and of M0INF.

```{r}
one_day_send_receive <- dbSendQuery(db_con,
  "SELECT * FROM spots
   WHERE (\"CallSign\" = 'M0INF' OR \"Reporter\" = 'M0INF')
   AND wd_time >= '2021-01-03T00:00:00Z'
   AND wd_time <  '2021-01-04T00:00:00Z';")
```

```{r}
the_dat_send_rec <- dbFetch(one_day_send_receive)
dbClearResult(one_day_send_receive)
```

```{r}
the_dat_send_rec <- the_dat_send_rec %>%
  mutate(Direction = ifelse(Reporter == "M0INF",
                            "Rx by M0INF",
                            "Tx by M0INF"))
```

The graph is a bit crowded so the y-axis is on the $\log_{10}$ scale.

```{r fig.height=5, fig.width=6, dpi = 300}
the_dat_send_rec %>%
  ggplot(aes(wd_time, distance, colour = Direction)) +
  geom_point(alpha = 0.5, size = 1) +
  stat_smooth(geom="line",
              alpha = 0.5, size = 0.8,
              method = "gam",
              formula = y ~ s(x, k = 10)) +
  labs(x = "Time", y = "Distance (km)",
       title = "WSPR reports",
       subtitle = "3 Jan 2021, 20m band") +
  scale_y_continuous(trans = "log10") +
  scale_x_datetime(date_breaks = "2 hour", date_labels = "%H") +
  theme(legend.position="bottom")
```

Which of those reports are over 5000 km?

```{r}
the_dat_send_rec %>%
  filter(distance > 5000) %>%
  group_by(CallSign, Reporter, distance) %>%
  summarise(Spots = n()) %>%
  arrange(desc(Spots)) %>%
  kable()
```

And which are under 100 km?

```{r}
the_dat_send_rec %>%
  filter(distance < 100) %>%
  group_by(CallSign, Reporter, distance) %>%
  summarise(Spots = n()) %>%
  arrange(desc(Spots)) %>%
  kable()
```
